---
title: "P8105 HW2"
author: "Yijin Serena Wang"
date: "`r Sys.Date()`"
output: github_document
---

```{r, message = FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)
```
## Problem 1
```{r}
subway_data <- read_csv("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  # drops last 9 columns
  select(-(last_col(offset = 8):last_col())) %>%
  # drop columns by name
  select(-Division, -Staffing, -`Staff Hours`) %>%
  clean_names() %>%
  mutate(entry=ifelse(entry=="YES", TRUE, FALSE))

```
This data set contains information about stations in NYC subway system. Each station has its name, line and routes associated with it. It also has information about each entrance of a station. For each entrance, there is information about entry type, entrance type (stairs/elevator/easement), whether it has vending machines and ADA compliance. When cleaning the data, I dropped the last 9 columns about more granular data for each entrance, and some other columns by instruction. Then I renamed all column names in camel case. The resulting dataset has `r ncol(subway_data)` columns and `r nrow(subway_data)` rows. It has a lot of repetitive information for each station across rows. It could be further cleaned. `route` variables could stay in a column rather than multiple columns with NAs.

```{r}
subway_data %>% 
  select(station_name, line) %>% 
  distinct
```
There are 465 distinct stations. 
```{r}
subway_data %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

84 of them are ADA compliant.

```{r}
subway_data %>%
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```
Approximately 3.78% of the station entrancs do not have vending machines.

```{r reformat}
cleaned_subway_data <- subway_data %>%
  mutate_at(vars(starts_with("route")), as.character) %>%
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE
  )
```

```{r}
cleaned_subway_data %>%
  filter(route_name == "A") %>% 
  select(station_name, line) %>% 
  distinct
```
There are 60 distinct stations that serve the A train. 

```{r}
cleaned_subway_data %>%
  filter(route_name == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```
17 of them are ADA compliant. 

## Problem 2
```{r Mr Trash Data}
# load and clean mr trash wheel data 
mr_trash_wheel_data <- read_excel(path = "./Data/Trash Wheel Collection Data.xlsx",
                                  sheet = "Mr. Trash Wheel",
                                  # skip pictures in row 1
                                  skip = 1) %>%
  # drops last two empty columns
  select(-(last_col(offset = 1):last_col())) %>%
  # clean names
  clean_names() %>%
  # filter data that are not specific to a dumpster
  filter(!is.na(dumpster)) %>%
  # round sports_balls column
  mutate(
    # round sports_balls column
    sports_balls = as.integer(sports_balls),
    # convert characters to numeric values for combining data later on
    year = as.numeric(year),
    # add extra column for combining data later on
    trash_wheel = "Mr.")
```

```{r Professor Trash Data}
# load and clean professor trash wheel data 
prof_trash_wheel_data <- read_excel(path = "./Data/Trash Wheel Collection Data.xlsx",
                                    sheet = "Professor Trash Wheel",
                                    # skip pictures in row 1
                                    skip = 1) %>%
  # clean names
  clean_names() %>%
  # filter data that are not specific to a dumpster
  filter(!is.na(dumpster)) %>%
  # add extra column for combining data later on
  mutate(trash_wheel = "Professor")
```
```{r combine}
# combine two data sets
results <- mr_trash_wheel_data %>%
  bind_rows(prof_trash_wheel_data)
```
Each data sets includes information about different types of trash in dumpsters that a trash wheel collected. There are two trash wheel: Mr.Trash Wheel and Professor Trash Wheel. The information includes specific date, weights, volumn and number of homes powered of a dumpster, and the number of each type of trash in it. These types include plastic bottles, polystyrene, cigarette butts, glass bottles, grocery bags, chip bags and sports balls. The combined `results` dataset has `r nrow(results)` rows and `r ncol(results)` columns of the information above. 
```{r}
results %>%
  filter(trash_wheel=="Professor") %>%
  pull(weight_tons) %>%
  sum
```   
The total weight of trash collected by Professor Trash Wheel is 190.12 tons. 

```{r}
results %>%
  filter(trash_wheel=="Mr." & results$year==2020) %>%
  pull(sports_balls) %>%
  sum
  
```

In 2020, the total number of sports balls collected by Mr. Trash Wheel is 856. 

## Problem 3
```{r pols-month}
# load and clean pols-month data
pols_month <- read_csv(file = "./Data/fivethirtyeight_datasets/pols-month.csv") %>%
  # break date column into month, day, year columns
  separate(col = mon,
           sep = "-",
           into = c("year", "month", "day")) %>%
  mutate(
    # convert month numbers to month names
    month = lubridate::month(as.integer(month),
                             label = TRUE,
                             abbr = FALSE),
    # convert characters to numeric data for merging later on
    year = as.integer(year)
  ) %>%
  # create president column from prez_dem and prez_gop
  # pivot two columns names into the president column
  pivot_longer(
    cols = c("prez_dem", "prez_gop"),
    names_to = "president",
    # keep dem and gop from columns names
    names_prefix = "prez_"
  )  %>%
  # only keep the rows that has value of 1 from previous step
  # there are 2s in prez_gop column. It might be a mistake but
  # the corresponding observations still should have "gop" value.
  filter(value >= 1) %>%
  # drop value column from pivot and day column
  select(-value,-day)
```

```{r snp}
#load and clean snp data
snp <- read_csv(file = "./Data/fivethirtyeight_datasets/snp.csv") %>%
  # break date column into month, day, year columns
  separate(col = date,
           sep = "/",
           into = c("month", "day", "year")) %>%
  mutate(
    # convert month numbers to month names
    month = lubridate::month(as.integer(month),
                             label = TRUE,
                             abbr = FALSE),
    # convert 2-digit year number to 4-digit year
    year = lubridate::year(
      # parse 2-digit year number to 4-digit year, with cutoff 15
      lubridate::parse_date_time2(year, "%y", cutoff_2000 = 15))
  ) %>%
  # arrange/order data by year and month
  arrange(year, month) %>%
  # make year and month as leading columns
  relocate(year, month, everything()) %>%
  # remove day column for merging later on
  select(-day)

```


```{r unemployment}
# load and clean unemployment data
unemployment <- read_csv(file = "./Data/fivethirtyeight_datasets/unemployment.csv") %>%
  # create a row for each month of a year and create a new column
  # for unemployment rate
  pivot_longer(-Year,
               names_to = "month",
               values_to = "unemployment_rate") %>%
  # convert month abbreviations to month names
  # match abbreviation to numbers, and then use numbers/indices to find
  # corresponding month name
  mutate(month = month.name[match(month, month.abb)]) %>%
  # rename Year to year
  rename(year = Year)
```

```{r merging}
# combine datasets
results <- pols_month %>%
  left_join(snp, by = c("year", "month"))  %>%
  left_join(unemployment, by = c("year", "month"))
  
```
The `pols_month` data set keeps track of national politicians' party for each month in 1947-2015. The `snp` data set contains closing prices of S&P stock index for each month in 1950-2015. The `unemployment` data set has unemployment rate for each month in 1948-2015. The combined dataset has `r ncol(results)` columns and `r nrow(results)` rows. The range of years is `r range(results$year)[1]` to `r range(results$year)[2]`. The closing prices are in `close` column. Unemployment rates are in `unemployment_rate` column. Counts of politicians in Democratic/Republican parties are in columns specified for each political level and each party. 